---
title: "R Data Structure Note"
author: "RuonanZhang"
date: "July 30, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Vectors

Vectors in R comes with `atomic vector` and `list`. Both have: 1) Type 2) Length 3)Attributes. Element of atomic vector must have same type while list doesn't. 

#### Atomic vectors
**`is.vector()` only return TRUE when the object is a vector with no attribute except names. **
```{r}
a = c(1,2,3) #a is an atomic vector
is.vector(a)
dim(a) = 3 #Adding dim attribute to a
is.vector(a)
```

Atomic vector has four common types.

1. Logical
2. integer
3. double (numeric)
4. charactor

```{r}
dbl_var = c(1, 2.5, 4.5)
# With the L suffix, you get an integer rather than a double, integer cosume less memory
int_var = c(1L, 6L, 10L) 
# Use TRUE and FALSE (or T and F) to create logical vectors
log_var = c(TRUE, FALSE, T, F)
chr_var = c("these are", "some strings")
```

The following are checking functions for each type
```{r}
is.character(chr_var)
is.double(dbl_var)
is.integer(int_var)
is.logical(log_var)
sapply(list(chr_var, dbl_var, int_var, log_var), is.atomic)
```

Be careful of `is.numeric()` command, it return TRUE for both int and double.
```{r}
is.numeric(int_var)
is.numeric(dbl_var)
```


What about NA?
```{r}
NA_test_dbl = c(NA, 2,3,4)
NA_test_chr = c(NA, 'b','c','d')
sapply(NA_test_dbl, class)
```
```{r}
sapply(NA_test_chr, class)
```
They are converted into common types by default

##### Coercion Sequence

Types from least to most flexible are: logical, integer, double, and character.

It means: double are more easy to coerce to character. Charactor would not be coerced to double by default.
```{r}
str(c("a", 1))
```

#### Lists

List elements can be of any type.
```{r}
x = list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

list() could combine two list and keep their struction. c() would flatten the list, which is always used as a method to `join` list.
```{r}
x <- list(list(1, 2), list(3, 4))
y <- c(list(1, 2), list(3, 4))
str(x)
str(y)
```

### Attributes

Remembered that both atomic vector and list have attributes. All objects can have arbitrary additional attributes, used to store metadata about the object. Attributes can be thought of as a named list (with unique names). Attributes can be accessed individually with `attr()` or all at once (as a list) with `attributes()`

```{r}
y = 1:10
attr(y, "my_attribute") = "This is a vector" # we just added a metadata for y
attr(y, "name") = "Name is y" # we just added a metadata for y
attributes(y)
```

```{r}
y
```

Most attribute would be lost when modifying the object.
```{r}
attributes(sum(y))
```

The only attributes not lost are the three most important:

**Names**, a character vector giving each element a name, described in names.

```{r}
x = c(a = 1, b = 2, c = 3)
str(x)

#The following also works
#x = 1:3
#names(x) = c("a", "b", "c")

#x = setNames(1:3, c("a", "b", "c"))
```


**Dimensions**, used to turn vectors into matrices and arrays, described in matrices and arrays.

**Class**, used to implement the S3 object system.

Those three attribute are not returned by `attributes()`, has to use their own functions.
```{r}
names(y) #y does not have name
dim(y) #y does not have dimension
class(y) 
class(sum(y)) 
```

#### Factors
A factor is a vector that can contain only predefined values, and is used to **store categorical data**. Factors are built on top of `integer` vectors using **two attributes**: the class, "factor", which makes them behave differently from regular integer vectors, and the levels, which defines the set of allowed values.

*Simple version: a factor is a integer vector with two special attributes.*

```{r}
x = factor(c("a", "b", "b", "a"))
class(x)
levels(x)
```

Why use factor: make it obvious when some level has no observations
```{r}
sex_char = c("m", "m", "m")
sex_factor = factor(sex_char, levels = c("m", "f"))

table(sex_char)
table(sex_factor)
```

Why is numerical value often converted to factor when read-in files?
```{r}
z = read.csv(text = "value\n12\n1\n.\n9")
typeof(z$value)
class(z$value) #Z$value is coerced into factor because missing value 
```

```{r}
as.double(z$value) #converting to double type would cause problem, the levels are used
```

Avoid this by using NA check when read in data.

