---
title: "Functional Programming R Notes"
author: "Ruonan Zhang"
date: "June 24, 2019"
output: html_document
---

This note does not concentrate on a systematic summary of functional programming. It list some application and examples that could often happens in regular data analysis. 


1. Dynamic Lookup & Lazy evaluation

R function is using dynamic lookup, which could make the function not self-contained.
```{r}
f = function() x
x = 15 #x is assigned after function is created, this should be avoided.
f()
```

Here's an example of lazy evaluation. When h is being called, it first create d, and b is evaluated until called. This should be avoided unless necessary.
```{r}
h = function(a = 1, b = d) { #d is not defined when creating the input
  d = (a + 1) ^ 2
  c(a, b)
}

h()
```

---

2. Function list

Get results/compare run times for different functions
```{r}
compute_mean <- list(
  base = function(x) mean(x),
  sum = function(x) sum(x) / length(x),
  manual = function(x) {
    total <- 0
    n <- length(x)
    for (i in seq_along(x)) {
      total <- total + x[i] / n
    }
    total
  }
)
```
Function could be an input of function, functions created by functions are called `Closure`, **because they enclose the environment of the parent function and can access all its variables**.

The anonymous function below creates closure. 
```{r}
x = runif(1e5)
lapply(compute_mean, function(f) system.time(f(x)))
```
Or use `...` as the notation of flexible inputs
```{r}
call_fun <- function(f, ...) f(...)
lapply(compute_mean, call_fun, x)
```

Function list could also be used in data description.

```{r}
x = c(1:10, NA)

funs = list(
  sum    = sum,
  mean   = mean,
  median = median
)
lapply(funs, function(f) f(x, na.rm = TRUE))
```

Another trick is to use `each`
```{r}
#Check out the each function in plyr
summaries = plyr::each(mean, sd, median)
summaries(1:10)
```



---

3. Infix function

Sometimes creating infix funcion could make code looks prettier. Infix function should be enclosed between %%. And illegal name could be used between ``.

```{r}
`%+%` = function(a, b) paste0(a, b)
"new" %+% " string"
```

The following application returns default value for a function calls incase it return's null. **This function could be very usefull when building Shiny apps.**

```{r}
`%||%` = function(a, b) if (!is.null(a)) a else b

function_that_might_return_null = function() NULL

function_that_might_return_null() %||% 'default value'
```

---

4. Replacement functions

Take a look at the error return of following code. What is "mean<-" ?
```{r error=TRUE}
x = 1:10
mean(x) = 999
```

A replacement function example is as follow. Replacement function could change object inplace. And it need to be named with a "<-"
```{r}
`second<-` = function(x, value) {
  x[2] = value
  x
}
x = 1:10
second(x) = 5
x
```

---

5. How to loop vectors

Using vectors to store element like this is slow. Because each time you extend the vector, R has to copy all of the existing elements.
```{r}
xs = runif(1e3)
res = c()
for (x in xs) {
  res = c(res, sqrt(x))
}
```

Instead:

```{r}
res = numeric(length(xs)) #Create empty vector and change in place
for (i in seq_along(xs)) { #seq_along is looping the vector by index
  res[i] = sqrt(xs[i])
}
```

---

6. Upgrading `sapply` to `vapply`.

vapply could be used to limit the output of sapply.

```{r}
sapply(list(), is.numeric)
vapply(list(), is.numeric, logical(1))
```

---

7. Special conditions that `lapply` or `sapply` does not work

When there are two or more list need to be looped through, try `Map`
```{r}
xs = replicate(5, runif(10), simplify = FALSE)
ws = replicate(5, rpois(10, 5) + 1, simplify = FALSE)

#This is the application of weighted mean of lapply
unlist(lapply(seq_along(xs), function(i) {
  weighted.mean(xs[[i]], ws[[i]])
}))

#Application in Map
unlist(Map(weighted.mean, xs, ws))
```

When the list is ragged group.
```{r}
pulse = round(rnorm(22, 70, 10 / 3)) + rep(c(0, 5), c(10, 12)) # All data in ragged group
group = rep(c("A", "B"), c(10, 12)) #Create index to separate the group

tapply(X = pulse, INDEX = group, FUN = length)
```

---

8. Merging several dataset with `Reduce`.

See following code for how `reduce` work.
```{r}
Reduce(`+`, 1:3) # -> ((1 + 2) + 3)
Reduce(sum, 1:3) # -> sum(sum(1, 2), 3)
```

Use `reduce` with `merge`
```{r}
a = data.frame(index = c('a','b','c'), val.a = c(1,2,3))
b = data.frame(index = c('a','b','c'), val.b = c(1,3,2))
c = data.frame(index = c('a','b','c'), val.c = c(2,1,3))
d = data.frame(index = c('a','b','c'), val.d = c(3,1,2))

Reduce(function(x,y) merge(x, y, by = 'index', all = TRUE), 
       list(a,b,c,d))

```

---

9. Use Memoization

See details below, memoization could be useful when building shiny app that would often retrieve data directly from a database that do not change regularly in a shiny session. 

https://www.inwt-statistics.com/read-blog/optimize-your-r-code-using-memoization.html

```{r}
#general idea of memoise is as follow
memoise = function(fun) {
  memory = list()
  function(x) {
    valueName = as.character(x)
    if (!is.null(memory[[valueName]])) return(memory[[valueName]])
    res = fun(x)
    memory[[valueName]] <<- res
    res
  }
}
```

---

10. Compose functions
```{r}
sapply(mtcars, function(x) length(unique(x)))
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(pryr)
sapply(mtcars, compose(length, unique))
```
or 

```{r}
library(magrittr)
sapply(mtcars, function(x) x %>% unique() %>% length()) #needs to define an extra anonymous function
```
---

11. Dealing with empty input in RShiny

Use `try` to assign value when it's possible that input doesn't exist
```{r}
x = 3
try(x > mean, silent = TRUE) #The error is surpressed
```


12. Use `quote` and `subs` (or `substitude`) when assigning name

13. Play with function calls

```{r}
mod = lm(mpg ~ wt, data = mtcars) #mod is a call
update(mod, formula = . ~ . + cyl) #update could change the element inside that call
```






